---
title: "Reference for 'tidyr' package"
author: 
date: "November 2, 2015"
output: 
  html_document: 
    keep_md: yes
---

  > *__tidy data__: every column in the data frame represents a variable and every row represents an observation --- also referred to as __long format__*
  
  
## Overview

* `gather()` and `spread()` convert between long and wide data
* `separate()` splits a single column into servaral variables and is commonly used in conjunction with `gather()`

```{r results='hide', message=FALSE, warning=FALSE}
library(tidyr); library(dplyr)

set.seed(1)
tidyr.ex <- data.frame(
  participant = c("p1", "p2", "p3", "p4", "p5", "p6"), 
  info = c("g1m", "g1m", "g1f", "g2m", "g2m", "g2m"),
  day1score = rnorm(n = 6, mean = 80, sd = 15), 
  day2score = rnorm(n = 6, mean = 88, sd = 8)
)
```

```{r}
print(tidyr.ex)
```

### gather: convert wide data to long data

* takes columns and gathers them into new variable
* `gather(df, newVar1, newVar2, vector1, vector2)`

```{r}
tidyr.ex %>%
  gather(day, score, c(day1score, day2score))
```


### spread: convert long data to wide data

* takes different levels of a factor and spreads them out into different columns
* `spread(df, var1, var2)`

```{r}
# returns back original data frame
tidyr.ex %>%
  gather(day, score, c(day1score, day2score)) %>%
  spread(day, score)
```


### separate: separates one column into many

* takes values in a column and creates new columns
* `separate(df, col, into, sep)`

```{r}
# sep can be numeric (length) or character
tidyr.ex %>%
  gather(day, score, c(day1score, day2score)) %>%
  separate(col = info, into = c("group", "gender"), sep = 2)
```


### unite: opposite of separate

* not used frequently
* `unit(df, newVarName, col1, col2)`

```{r}
# default combines with "_"
tidyr.ex %>%
  gather(day, score, c(day1score, day2score)) %>%
  separate(col = info, into = c("group", "gender"), sep = 2) %>%
  unite(infoAgain, group, gender)
```

